A second example of extremely reproducible enrichment analysis

### 2023-06-30 example2.Rmd
code edits

---
title: "An example of extremely reproducible enrichment analysis"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

@@

For this guide I will be using bulk RNA-seq data from a previous study, which is
deposited at NCBI GEO and SRA under accession numbers: GSE55123 and SRP038101 (Lund et al, 2014).
The experiment is designed to investigate the effect of Azacitidine treatment on AML3 cells.

The raw data have been processed by the DEE2 project, and the summary gene expression counts are
available at the dee2.io website, and programmatically with the getDEE2 bioconductor package
(Ziemann et al, 2019).

The gene counts have also been deposited to the `/data` folder in the `example.Rdata` file in case
the DEE2 resource becomes unavailable.
To import it, use the command: `load("data/example.Rdata")`

Alternatively, you could fetch data from another resource like NCBI GEO, Zenodo or from the host
storage drive.

@@

mdat <- getDEE2Metadata("hsapiens")

# get sample sheet
ss <- subset(mdat,SRP_accession=="SRP038101")

# fetch the whole set of RNA-seq data
x <- getDEE2("hsapiens",ss$SRR_accession , metadata=mdat, legacy=TRUE)
mx <- x$GeneCounts
rownames(mx) <- paste(rownames(mx),x$GeneInfo$GeneSymbol)
dim(mx)

# aza no filtering
ss$trt <- grepl("Treated",ss$Experiment_title)

ss %>%
  kbl(caption="sample sheet for Aza treatment in AML3 cells") %>%
  kable_paper("hover", full_width = F)

@@

cols <- as.character(grepl("Treated",ss$Experiment_title))
cols <- gsub("FALSE","lightblue",cols)
cols <- gsub("TRUE","pink",cols)
plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(XMIN,XMAX),ylim=c(YMIN,YMAX),
  pch=19,cex=2,col=cols, main="MDS plot")
text(cmdscale(dist(t(mx))), labels=colnames(mx) )

@@

plot(log10(def$baseMean) ,def$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  main="5-azacitidine treatment in AML3 cells",
  xlab="log10(basemean)",ylab="log2(fold change)")

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | via DEE2 at accession SRP038101 (human) |
| Other parameters | Min gene set size of 10 |

@@

One of the quirky things I learned about clusterprofiler is that
not all genes specified by the user as background are included in
the enrichmentcalculation.
It appears as though the genes not included in the annotation set are
discarded.
I believe this is a subtle but significant problem.
To fix it is relatively easy from the user's point of view.
Just create a new gene set that consists of all detected genes and

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | FGSEA (check version at foot of report) |
| Statistical test used | Kolmogorov-Smirnov test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | via DEE2 at accession SRP038101 (human) |
| Other parameters | Min gene set size of 10 |

>>>>

---
title: "A second example of extremely reproducible enrichment analysis"
author: "Mark Ziemann & Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 5
    fig_height: 5
theme: cosmo
---

@@

For this guide I will be using bulk RNA-seq data from a previous study, which is
deposited at NCBI GEO with [accession GSE93236](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE93236).

The experiment is designed to investigate the effect of knowckdown of the histone methyltransferase set7.

We will be fetching the gene expression count matrix directly from NCBI GEO for downstream analysis.
Link: https://ftp.ncbi.nlm.nih.gov/geo/series/GSE93nnn/GSE93236/suppl/GSE93236_Keating_Set7_KD_counts.tsv.gz

@@

download.file("https://ftp.ncbi.nlm.nih.gov/geo/series/GSE93nnn/GSE93236/suppl/GSE93236_Keating_Set7_KD_counts.tsv.gz",
  destfile="GSE93236_Keating_Set7_KD_counts.tsv.gz")

x <- read.table("GSE93236_Keating_Set7_KD_counts.tsv.gz")

dim(x)

# there is a row we need to remove
mx <- x[rownames(x)!="NOTINBED",]

head(mx)

dim(mx)

The row names contains an underscore that we need to remove.

```{r,geneid_fix}

rownames(mx) <- sub("_"," ",rownames(mx))

head(mx)

Curate sample sheet from the colnames.

```{r,samplesheet1}

ss <- as.data.frame(colnames(x))
ss$trt <- as.numeric(grepl("Set7KD",colnames(x)))

ss %>%
  kbl(caption="sample sheet for Set7KD experiment") %>%
  kable_paper("hover", full_width = F)

@@

cols <- as.character(ss$trt)
cols <- gsub("0","lightblue",cols)
cols <- gsub("1","pink",cols)
plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  xlim=c(XMIN,XMAX),ylim=c(YMIN,YMAX),
  pch=19,cex=2,col=cols, main="MDS plot")
text(cmdscale(dist(t(mx))), labels=colnames(mx) )

@@

plot(log10(def$baseMean) ,def$log2FoldChange,
  cex=0.6,pch=19,col="darkgray",
  main="Set7KD in endothelial cells",
  xlab="log10(basemean)",ylab="log2(fold change)")

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | ClusterProfiler (check version at foot of report) |
| Statistical test used | hypergeometric test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | GEO under accession GSE93236 |
| Other parameters | Min gene set size of 10 |

@@

One of the quirky things I learned about clusterprofiler is that
not all genes specified by the user as background are included in
the enrichmentcalculation.
It appears as though the genes not included in the annotation set are
discarded.
I believe this is a subtle but significant problem.
To fix it is relatively easy from the user's point of view.
Just create a new gene set that consists of all detected genes and add it
to the main gene set data frame.

@@

| Reporting criteria | Method/resource used |
| --- | --- |
| Origin of gene sets | Reactome (2023-03-06) |
| Tool used | FGSEA (check version at foot of report) |
| Statistical test used | Kolmogorov-Smirnov test |
| P-value correction for multiple comparisons | FDR method |
| Background list | Genes with >=10 reads per sample on average across all samples |
| Gene Selection Criteria | DESeq2 FDR<0.05 |
| ORA directionality | Separate tests for up- and down-regulation |
| Data availability | GEO under accession GSE93236 |
| Other parameters | Min gene set size of 10 |
